A Solver for Generalised Parikh Images of Regular Languages

### 2021-10-11 abstract/main.tex
conclusions

Parikh images are a central notion in automata theory and context-free grammars,
with wide-ranging applications, including the solving
of string constraints in automata-based solvers such as \Ostrich{}
\cite{ostrich}. In this work, we focus on the efficient computation
of Parikh images, or points within the Parikh image, of finite-state
automata. While it is possible to compute the Parikh image of an automaton
by adapting the method for context-free grammars
 of~\cite{generate-parikh-image}, this method
produces clauses with many existentially quantified variables which are costly
to eliminate. This makes many real-world applications intractable. Furthermore,
applications like Unicode require automata with symbolic labels to handle their
large alphabets, increasing the number of variables even more.
Finally, automata-based string solvers compute Parikh images on products of
automata, derived from conjunctions of string constraints. Using
\cite{generate-parikh-image} this would require the up-front computation of the
product before its Parikh image, running the risk of an exponential blow-up.

@@

An example is computing the length of a string, which is expressed in terms of
the Parikh map by summing the individual character counts of the vector: $h':\:
(\mathbb{Z}^\Sigma, +, \vec{0}) \to (\mathbb{Z}, +, 0) = \vec{x} \to \sum_{i \in
\Sigma} x_i$. More generally, this approach extends to multiple counters to
produce Parikh automata with one such counter per symbol in the language, as
used in implementations of many string
constraints~\cite{path-feasibility-strings}.

\section{Lazy Computation of Parikh Images for Regular Languages}
The calculus is based on four principles: linear equations to preserve flow
through the automaton, case-splitting when there are multiple possible paths a
run can take, propagation of connectedness constraints, and materialisation of
products. Notably, ensuring connectedness of a run in the presence of cycles is
the computationally most difficult constraint to enforce, as the flow equations
of any loop will self-balance.

@@

>>>>

Parikh images are a central notion in automata and language theory,
with wide-ranging applications, including the solving
of string constraints in automata-based solvers such as \Ostrich{}~\cite{ostrich}. In this work, we focus on the efficient computation
of Parikh images, or points within the Parikh image, of finite-state
automata. While it is possible to compute the Parikh image of an automaton
by adapting the method for context-free grammars
 of~\cite{generate-parikh-image}, this method
produces clauses with many existentially quantified variables which are costly
to eliminate. This makes many real-world applications intractable. Furthermore,
applications like Unicode require automata with symbolic labels to handle their
large alphabets, increasing the number of variables even more.
Finally, automata-based string solvers compute Parikh images on products of
automata, derived from conjunctions of string constraints. Using
\cite{generate-parikh-image} this would require the up-front computation of the
product before its Parikh image, running the risk of an exponential blow-up.

@@

An example is computing the length of a string, which is expressed in terms of
the Parikh map by summing the individual character counts of the vector: $h':\:
(\mathbb{Z}^\Sigma, +, \vec{0}) \to (\mathbb{Z}, +, 0) = \vec{x} \to \sum_{i \in
\Sigma} x_i$. More generally, this approach extends to multiple counters to
produce Parikh automata with one such counter per symbol in the language, as
used in some string
constraint solvers~\cite{path-feasibility-strings}.

\section{Lazy Computation of Parikh Images for Regular Languages}
In the scope of this abstract, we introduce our calculus for computing
Parikh images through an example.
The calculus is based on four principles: linear equations to preserve flow
through the automaton, case-splitting when there are multiple possible paths a
run can take, propagation of connectedness constraints, and materialisation of
products. Notably, ensuring connectedness of a run in the presence of cycles is
the computationally most difficult constraint to enforce, as the flow equations
of any loop will self-balance.

@@

\section{Conclusions}

We are currently in the process of finishing the first version of our
Parikh image tool. As an evaluation, we will use the tool to compute
length abstractions of regular expression constraints in string
solving, and to handle the automata augmented with counters introduced
in \cite{path-feasibility-strings}. We will compare with the approach
proposed in \cite{generate-parikh-image}, and with a translation to a
model checking problem, which can then be solved using (more general)
symbolic and abstraction-based model checkers.
